\documentclass{article}
\usepackage{preamble}

\begin{document}
\header{Filemon Mateus}{CS 6150}{Graduate Algorithms}{Homework \# 5}{\today}
\begin{enumerate}[leftmargin={*}, font={\bf}, label={\arabic*.}, ref={\arabic*}]
  \item \label{qst:1}
    \begin{enumerate}
      \item \label{qst:1a}
        {\itshape Algorithm Description}

        We proceed with dynamic programming. Specifically, for any integer $0 \leq i \leq n$ and
        $0 \leq j \leq M$, we split the original problem into individual subproblem instances of
        the form $f[i][j]$, each representing a boolean value that is \textsc{True} \textit{iff}
        there exists a multiset of items $1, \ldots, i$ of size $j$. With a abuse of set notation
        we define:
        \[
          f[i][j] = 1 \iff \exists\ S \subseteq \{1, \ldots, i\} \ni \sum_{i \in S} a_i = j
        \]
        At iteration $i$, we discuss the progress made by our dynamic program in one of the
        following ways:
        \begin{enumerate}[label={(\arabic*)}, itemsep={0pt}]
          \item
            Item $i$ occurs at least once in $S$. In this case, we select a multiset of items $1,
            \ldots, i$ that has size exactly $j-a_i$. Here $f[i][j-a_i]$ indicates whether such
            multiset exists.
          \item
            Item $i$ does not occur in $S$. In this case, the solution maps to a multiset of the
            items $1, \ldots, i-1$ of size $j$. The answer of whether such multiset exists is found
            in $f[i-1][j]$.
        \end{enumerate}
        Either one of the above avenues yields a solution, so $f[i][j]$ is \textsc{True} if at least
        one of the decision possibilities above results in a solution. Furthermore, we can formalize
        these decision states as a dependency relation among the subproblems of $f$. Namely:
        \[
          f[i][j] = 
          \begin{cases}
            f[i-1][j] & \text{if}\ j < a_i \\
            \max\big\{f[i-1][j], f[i][j-a_i]\big\} & \text{otherwise}
          \end{cases}
        \]
        The base case follows from the definition of $f[i][j]$, i.e., $f[i][0] = 1$ for all $0 \leq
        i \leq n$ and $f[0][j] = 0$ for all $1 \leq j \leq M$. The answer we are after is $f[n][M]$.
        Since we do not know where it resides, we try all possible pairs $(i,j)$ and check if any of
        them evaluate to \textsc{True}---and this is achieved by taking the max across the intermediate
        answers to $f[n][M]$.

        Note this is a two-dimensional dynamic programming problem with table size $(n+1, M+1)$.
        Each table entry is a boolean value (\textsc{True/False}) capturing the inclusion (or
        exclusion) of item $i$ in the final multiset $S$. Each such decision incurs an $\O(1)$ work
        (deciding among the two previous precomputed subproblems which one is \textsc{True}) so the
        total running time is in the order of $\O(nM)$.

      \item \label{qst:1b}
        {\itshape Algorithm Pseudocode} \vspace{-\baselineskip}

        \begin{minipage}{\linewidth}
          \begin{algorithm}[H]
            \caption{$\textsc{Knapsack}(A, M)$}\label{alg:unbounded-knapsack}
            \begin{algorithmic}[1]
              \Require{A list of $n$ item sizes $A[1 \cdots n]$ together with an integer $M$---the size of the knapsack.}
              \Ensure{{\color{red!50!black}``\texttt{yes}''} \textit{iff} there exists a multiset $S$ of items of $A$ whose total size is equal to $M$; {\color{red!50!black}``\texttt{no}''} otherwise.}
              \State $f[i][0] \gets 1$ \textbf{for each} $i \in \{0 \cdots n\}$
              \State $f[0][j] \gets 0$ \textbf{for each} $j \in \{1 \cdots M\}$
              \For{$i \gets 1$ \textbf{to} $n$}
                \For{$j \gets 1$ \textbf{to} $M$}
                  \If{$j < A[i]$}.
                    \State $f[i][j] \gets f[i-1][j]$
                  \Else
                    \State $f[i][j] \gets \max\big\{f[i-1][j], f[i][j-A[i]]\big\}$
                  \EndIf
                \EndFor
              \EndFor
              \If{$f[n][M] = 1$}
                \State \Return {\color{red!50!black}``\texttt{yes}''}
              \Else
                \State \Return {\color{red!50!black}``\texttt{no}''}
              \EndIf 
            \end{algorithmic}
          \end{algorithm}
        \end{minipage}
    \end{enumerate}

  \newpage

  \item \label{qst:2}
    \begin{enumerate}
      \item \label{qst:2a}
        {\itshape Algorithm Description}

        We proceed with dynamic programming. Specifically, for any integer $0 \leq i \leq n$ and $0
        \leq j \leq M$, we split the original problem into individual subproblem instances of the
        form $f[i][j]$ denoting the maximum attainable value obtained using a knapsack of capacity
        $j$ and the first $i$ items $1, \ldots, i$. At iteration $i$, we face a decision problem;
        that is, either item size $a_i$ is needed to achieve the optimal solution, or it isn't needed.
        Since we want to maximize $\sum_{a_i \in S} v_i$, we select the most profitable result among
        the two following decision states:

        \begin{enumerate}[label={(\arabic*)}, itemsep={0pt}]
          \item
            Fill the knapsack with item $i$. In this case, we must select a subset of items $1, \ldots,
            i-1$ that have a combined size at most $j - a_i$. Assuming we do all the preceding $i-1$
            choices correctly, then we will get $f[i-1][j-a_i]$ value out of items $1, \ldots, i-1$,
            so the cumulative total will be $f[i-1][j-a_i] + v_i$.
          \item
            Don't fill the knapsack with item $i$, so we'll re-use the optimal solution for items $1,
            \ldots, i-1$ that have an aggregate size at most $j$. That answer rests in $f[i-1][j]$.
        \end{enumerate}

        We can formalize $(1)$ and $(2)$ as the following dependency relation among the subproblems
        of $f$:
        \[
          f[i][j] =
            \begin{cases}
              f[i-1][j] & \text{if}\ j < a_i \\
              \max\big\{f[i-1][j], f[i-1][j-a_i] + v_i\big\} & \text{otherwise}
            \end{cases}
        \]
        Note the initial conditions for this problem are $f[i][0] = 0$ for all $0 \leq i \leq n$ 
        and $f[0][j] = 0$ for all $0 \leq j \leq M$. The answer we seek is $f[n][M]$. Since we do
        not know which pair $(i,j)$ yields $f[n][M]$, we try all possible combinations of $i$'s and
        $j$'s and pick the maximal among them. The pseudocode for this is provided below and boils
        down to filling out a two dimensional table of size $(n+1, M+1)$. Each entry of the table
        incurs an $\O(1)$ work (deciding among the two previous precomputed subproblems which one
        is larger) so the total running time is $\O(nM)$.

      \item \label{qst:2b}
        {\itshape Algorithm Pseudocode} \vspace{-\baselineskip}

        \begin{minipage}{\linewidth}
          \begin{algorithm}[H]
            \caption{$\textsc{Knapsack}(A, V, M)$}\label{alg:bounded-knapsack}
            \begin{algorithmic}[1]
              \Require{A list of $n$ item sizes $A[1 \cdots n]$, its corresponding values $V[1 \cdots n]$, and an integer $M$---the size of the knapsack.}
              \Ensure{The maximum attainable value with a knapsack of size $M$.}
                \State $f[i][0] \gets 0$ \textbf{for each} $i \in \{0 \cdots n\}$
                \State $f[0][j] \gets 0$ \textbf{for each} $j \in \{0 \cdots M\}$
                \For{$i \gets 1$ \textbf{to} $n$}
                  \For{$j \gets 1$ \textbf{to} $M$}
                    \If{$j < A[i]$}
                      \State $f[i][j] \gets f[i-1][j]$
                    \Else
                      \State $f[i][j] \gets \max\big\{f[i-1][j], f[i-1][j-A[i]] + V[i]\big\}$
                    \EndIf
                  \EndFor
                \EndFor
                \State \Return $f[n][M]$
            \end{algorithmic}
          \end{algorithm}
        \end{minipage}
    \end{enumerate}

  \newpage

  \item \label{qst:3}
    \begin{enumerate}
      \item \label{qst:3a}
        {\itshape Algorithm Description}
        
        The maximum sum common subsequence (MSCS) problem has identical decision states to the longest
        common subsequence (LCS) problem discussed in class. So, without prizes for guessing, we proceed
        with dynamic programming. Specifically, for integers $0 \leq i \leq n$ and $0 \leq j \leq m$, we
        introduce individual subproblem instances of the form $f[i][j]$ representing the MSCS attainable
        from inputs $A[1 \cdots i]$ and $B[1 \cdots j]$. When computing $f[i]$[j], we take into account
        one of the following decision states:

        \begin{enumerate}[label={(\arabic*)}, itemsep={0pt}]
          \item
            $A[i] = B[j]$. Then clearly the MSCS is obtained by maximizing the sum of the subsequences
            $A[1 \cdots i-1]$ and $B[1 \cdots j-1]$, then adding $A[i]$ (or $B[j]$) to both subsequences.
            Notice, however, that there is a small caveat here, namely: $A[i]$ could be negative! Since
            we want to maximize the sum we add $A[i]$ only if $A[i]$ is positive. This should satisfiably
            handle negative numbers in $A[1 \cdots i]$ and $B[1 \cdots j]$ and make our dynamic program
            robust against edge cases of this nature.
          \item
            $A[i] \neq B[j]$. Then clearly the optimal subsequence yielding MSCS is located in the shorter
            version of $A[1 \cdots i]$ and $B[1 \cdots j]$. Since we do not know which one it is, we solve
            both $f[i-1][j]$ and $f[i][j-1]$ and pick the corresponding result that yields the maximal sum
            among them.
        \end{enumerate}

        Formalizing the aforementioned decision states as a dependency relation across the subproblems of
        $f$, we see that:
        \[ 
          f[i][j] = 
          \begin{cases}
            f[i-1][j-1] + \max\big\{A[i], 0\big\} & \text{if}\ A[i] = B[j] \\
            \max\big\{f[i][j-1], f[i-1][j]\big\} & \text{if}\ A[i] \neq B[j]
          \end{cases}
        \]
        The base case follows from the definition of $f[i][j]$, i.e., $f[i][0] = 0$ for all $0 \leq i \leq
        n$ and $f[0][j] = 0$ for all $0 \leq j \leq m$. We are after $f[n][m]$, but since we do not know
        which pairs of indices $(i, j)$ yield $f[n][m]$, we try all permissible choices of $i$'s and $j$'s
        and pick the maximal one among them.

        Due to the optimal subproblem structure similarity to LCS, we see that MSCS is also a two dimensional
        dynamic program of table size $(n+1, m+1)$. Retrieving the value of $f[i][j]$ at any given iteration
        takes $\O(1)$. There are $(n+1)(m+1)$ entries in the table, so we get a final, total running time of
        $\O(nm)$ as required! The pseudocode for this is provided below.

      \item \label{qst:3b}
        {\itshape Algorithm Pseudocode} \vspace{-\baselineskip}

        \begin{minipage}{\linewidth}
          \begin{algorithm}[H]
            \caption{$\textsc{Max-Sum-Common-Subsequence}(A, B)$}\label{alg:max-sum-commmon-subsequence}
            \begin{algorithmic}[1]
              \Require{An array $A[1 \cdots n]$ of $n$ numbers and an array $B[1 \cdots m]$ of $m$ numbers.}
              \Ensure{The maximum sum common subsequence of $A$ and $B$.}
              \State $f[i][0] \gets 0$ \textbf{for each} $i \in \{0 \cdots n\}$
              \State $f[0][j] \gets 0$ \textbf{for each} $j \in \{0 \cdots m\}$
              \For{$i \gets 1$ \textbf{to} $n$}
                \For{$j \gets 1$ \textbf{to} $m$}
                  \If{$A[i] = B[j]$}
                    \State $f[i][j] \gets f[i-1][j-1] + \max\big\{A[i], 0\big\}$
                  \Else
                  \State $f[i][j] \gets \max\big\{f[i][j-1], f[i-1][j]\big\}$
                  \EndIf
                \EndFor
              \EndFor
              \State \Return $f[n][m]$
            \end{algorithmic}
          \end{algorithm}
        \end{minipage}
    \end{enumerate}

  \newpage

  \item \label{qst:4}
    \begin{enumerate}
      \item \label{qst:4a}
        {\itshape Algorithm Description}

        We proceed with dynamic programming. Specifically, for any $1 \leq i \leq n$, we split the
        original problem into individual subproblem instances of the form $f[i]$ representing the
        length of the longest monotonically increasing subsequence in $A[1 \cdots i]$ that ends with
        $A[i]$.

        We compute the value of $f[i]$ by considering all of the $i-1$ elements in $A$ both previous
        to $A[i]$ and smaller than it, i.e., $\{j \in [1 \cdots i-1] \ni A[i] > A[j]\}$.  These are
        the potential candidates that $A$ could be appended to in an increasing subsequence. Since
        we are after the longest one, we choose the $f[j]$ with maximal value, and set $f[i] = 1 +
        f[j]$ (thereby effectively adding $A[i]$ to the end of the longest monotonically increasing
        subsequence possible). So we have:
        \[
          f[i] = 1 + \max\big\{f[j] \ni 1 \leq j < i \land A[i] > A[j]\big\}
        \]
        The base case rests on $f[1] = 1$. Note $f[i]$ depends on all of the elements before it, so
        when we create our dynamic programming table, we start with the base case $f[1]$ and then
        progressively fill it from left to right. Once we determine the values for all $f[i]$, we
        just select the one with maximum value, and the algorithm is complete. Recovering the actual
        subsequence is a near trivial task of keeping track of which $A[i]$'s are added to the
        sequence and then backtrack them. The pseudocode for this is given below and has a time
        complexity of $\O(n^2)$ incurred by the two \texttt{for} loops in Steps $6$ and $8$.

      \item \label{qst:4b}
        {\itshape Algorithm Pseudocode} \vspace{-\baselineskip}

        \begin{minipage}{\linewidth}
          \begin{algorithm}[H]
            \caption{$\textsc{Longest-Monotonically-Increasing-Subsequence}(A)$}\label{alg:longest-monotonically-increasing-subsequence}
            \begin{algorithmic}[1]
              \Require{An array $A[1 \cdots n]$ of $n$ distinct numbers.}
              \Ensure{The longest monotonically increasing subsequence $S$ of $A$ and its length $l$.}
              \State $l \gets 0$
              \State $k \gets 0$
              \State $S \gets \emptyset$
              \State $p[1 \cdots n] \gets 0$ \Comment{predecessors for each $i$'s -- useful for reconstructing $S$}
              \State
              \For{$i \gets 1$ \textbf{to} $n$}
                \State $f[i] \gets 1$
                \For{$j \gets 1$ \textbf{to} $i-1$}
                  \If{$A[i] > A[j]$ \textbf{and} $f[i] < f[j] + 1$}
                    \State $f[i] \gets f[j] + 1$
                    \State $p[i] \gets j$
                  \EndIf
                \EndFor
                \If{$l < f[i]$}
                  \State $l \gets f[i]$
                  \State $k \gets i$
                \EndIf
              \EndFor
              \State
              \Repeat
                \State $S \gets S \cup \{A[k]\}$
                \State $k \gets p[k]$
              \Until{$k = 0$}
              \State
              \State $S \gets \Call{Reverse}{S}$
              \State \Return $(S, l)$
            \end{algorithmic}
          \end{algorithm}
        \end{minipage}
    \end{enumerate}
\end{enumerate}
\end{document}
